Dependently Typed Languages in Statix

### 2023-01-16 paper-evcs/paper.tex
Swap paragraph and figure
Changes based on Aron's feedback v3

The base language that has been implemented is the Calculus of Constructions~\cite{Coquand_Huet_1988}. One extra feature was added that is not present in the Calculus of Constructions: let bindings. Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful. The concrete syntax~\cite{sdf3} of the language is available in figure \ref{fig:syntax}.

@@

\begin{lstlisting}
let f = \T: Type. \x: T. x;
f (T: Type -> Type) (\x: Type. x)
\end{lstlisting}

@@

We need to define multiple predicates that will be used later for type checking. First, the primary predicate is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from~\cite{krivine}, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $\bhr{\scope{s_1}{e_1}}{t}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|. The \verb|t| in this definition is the argument stack of the Krivine machine. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and the stack of arguments (of the Krivine machine state) and converts it to an expression by adding \verb|FnDestruct|s. Its signature is:
\begin{lstlisting}
rebuild : (scope * Expr) * list((scope * Expr)) -> (scope * Expr)
\end{lstlisting}

@@

\begin{figure}[h]
	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{t}{\scope{s_2}{e_2}}$}]{Beta head-reduction rules}

@@

		% Let rule
		\inferrule{ 
			\bhr 
			{ \scope{ \co{sPutSubst}(s, x, (s, e))}{b} } 
			{ t }
			{ \scope{ s' } { b' } }
		}{
			\bhr
			{ \scope{ s }{ \co{Let}(x, e, b) } }
			{ t }
			{ \scope { s' } { b' } }
		}

		% Var rule - NSubst
		\inferrule{ 
			\co{sGetName}(s, x) = \co{NSubst}(s_e, e) 
			\\ \bhr
			{\scope{s_e}{e}}
			{t}
			{\scope{s_{e'}}{e'}} 
		}{
			\bhr
			{\scope{s}{\co{Var}(x)}}
			{t}
			{\scope{s_{e'}}{e'}} 
		}

		% Var rule - NType
		\inferrule{ 
			\co{sGetName}(s, x) = \co{NType}(t) 
		}{ 
			\bhr
			{\scope{s}{\co{Var}(x)}}
			{t}
			{\co{rebuild}(s, \co{Var}(x), t)} 
		}

@@

		% FnConstruct rule - Args
		\inferrule{ 
			\bhr
			{\scope{\co{sPutSubst}(s, x, t)}{b}}
			{ts}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{\co{FnConstruct}(x, \_, b)}}
			{(t::ts)}
			{\scope{s'}{e'}} 
		}

		% FnDestruct rule
		\inferrule{ 
			\bhr
			{\scope{s}{f}}
			{(a::ts)}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{\co{FnDestruct}(f, a)}}
			{ts}
			{\scope{s'}{e'}} 
		}
	\end{mathpar}
	\caption{Rules for beta head reducing the Calculus of Constructions}
	\label{fig:beta-head-reduce-rules}
\end{figure}

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
\begin{enumerate}
	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1), t) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

@@

For example, in the following code it would only suggest expressions that can be booleans (including variables that are booleans):
\begin{lstlisting}
let f = \b: Bool. b;
f {{Expr}}
\end{lstlisting}

>>>>

The base language that has been implemented is the Calculus of Constructions~\cite{Coquand_Huet_1988}. One extra feature was added that is not present in the Calculus of Constructions: let bindings. Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful. The concrete syntax (written in SDF3~\cite{sdf3}) of the language is available in figure \ref{fig:syntax}.

@@

\begin{lstlisting}
let f = \T: Type. \x: T. x;
f (T: Type -> Type) (\y: Type. y)
\end{lstlisting}

@@

We need to define multiple predicates that will be used later for type checking. First, the primary predicate is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from~\cite{krivine}, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), ps) == (s2, e2)|. The \verb|p| in this definition is the argument stack of the Krivine machine. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and the stack of arguments (of the Krivine machine state) and converts it to an expression by adding \verb|FnDestruct|s. Its signature is:
\begin{lstlisting}
rebuild : (scope * Expr) * list((scope * Expr)) -> (scope * Expr)
\end{lstlisting}

@@

\begin{figure}[h]
	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$}]{Beta head-reduction rules}

@@

		% Let rule
		\inferrule{ 
			\bhr 
			{ \scope{ \co{sPutSubst}(s, x, (s, e))}{b} } 
			{ \overline{p} }
			{ \scope{ s' } { b' } }
		}{
			\bhr
			{ \scope{ s }{ \co{Let}(x, e, b) } }
			{ \overline{p} }
			{ \scope { s' } { b' } }
		}

		% Var rule - NSubst
		\inferrule{ 
			\co{sGetName}(s, x) = \co{NSubst}(s_e, e) 
			\\ \bhr
			{\scope{s_e}{e}}
			{\overline{p}}
			{\scope{s_{e'}}{e'}} 
		}{
			\bhr
			{\scope{s}{\co{Var}(x)}}
			{\overline{p}}
			{\scope{s_{e'}}{e'}} 
		}

		% Var rule - NType
		\inferrule{ 
			\co{sGetName}(s, x) = \co{NType}(t) 
		}{ 
			\bhr
			{\scope{s}{\co{Var}(x)}}
			{\overline{p}}
			{\co{rebuild}(s, \co{Var}(x), \overline{p})} 
		}

@@

		% FnConstruct rule - Args
		\inferrule{ 
			\bhr
			{\scope{\co{sPutSubst}(s, x, p)}{b}}
			{\overline{p}}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{\co{FnConstruct}(x, \_, b)}}
			{(p::\overline{p})}
			{\scope{s'}{e'}} 
		}

		% FnDestruct rule
		\inferrule{ 
			\bhr
			{\scope{s}{f}}
			{(a::\overline{p})}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{\co{FnDestruct}(f, a)}}
			{\overline{p}}
			{\scope{s'}{e'}} 
		}
	\end{mathpar}
	\caption{Rules for beta head reducing the Calculus of Constructions}
	\label{fig:beta-head-reduce-rules}
\end{figure}

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
\begin{enumerate}
	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $\bhr{\scope{s1}{e1}}{\overline{p}}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1), ps) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

@@

For example, in the following code it would only suggest expressions that can be booleans:
\begin{lstlisting}
let f = \b: Bool. b;
f [[Expr]]
\end{lstlisting}

This would return the following suggestions: (Note that \verb|f| and \verb|Type| are not suggested, since they cannot have type boolean)
\begin{itemize}
	\item \verb|[[Expr]] [[Expr]]|
	\item \verb|let [[ID]] = [[Expr]]; [[Expr]]|
	\item \verb|true|
	\item \verb|false|
	\item \verb|if [[Expr]] then [[Expr]] else [[Expr]] end|
\end{itemize}
